A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-09 latex/main.tex
Pass through intro incorporating RIPPLES

Here we present sc2ts, a new method to infer ARGs for SARS-CoV-2  
that incrementally adds samples over time in a similar manner 
to UShER, but also automatically detects and incorporates recombination
in real-time. Based on the Viridian dataset, a freely available  
resource purged of various systematic errors and artifacts \cite{Hunt2024},
we infer an ARG with 2.48 million whole genome sequences.
We show that this ARG is highly congruent with the UShER phylogeny built 
from the same dataset as well as the Pango lineage naming system\cite{Rambaut2020},
and also accurately captures evolutionary features such as mutational 
spectra\cite{Bloom2023} and deletions\cite{Li2023indels}.
In addition, we show that sc2ts automatically and accurately detects details about
known recombinant lineages that were painstakingly collated in a community 
process, and provides novel insights into the relationships between these
recombinant lineages. By fully incorporating these and other novel
recombination events, the resulting data structure---representing
genealogical relationships, point mutations, deletions, and 
recombination---provides a basis for the study of SARS-CoV-2
that systematically accounts for all of these evolutionary forces.
This ARG, coupled with the rich suite of supporting software
in the tskit\cite{Ralph2020} and VCF-Zarr~\cite{Czech2025} ecosystems,
enables pandemic-scale analysis on a laptop. 
% Dunno about this! Feels important to finish on a wider note, helping
% the reader understand the significance of all of this?
These resources
will help us to learn from the vast troves of data collected during the
pandemic, and to prepare for the next.

>>>>

Here we present sc2ts, a new method to infer ARGs for SARS-CoV-2  
that incrementally adds samples over time in a similar manner 
to UShER, but also automatically detects and incorporates recombination
in real-time. Based on the Viridian dataset, a freely available  
resource purged of various systematic errors and artifacts \cite{Hunt2024},
we infer an ARG with 2.48 million whole genome sequences.
We show that this ARG is highly congruent with the UShER phylogeny built 
from the same dataset as well as the Pango lineage naming system\cite{Rambaut2020},
and also accurately captures evolutionary features such as mutational 
spectra\cite{Bloom2023} and deletions\cite{Li2023indels}.
We show that sc2ts automatically and accurately detects details about
known recombinant lineages that were painstakingly collated in a community 
process, and provides novel insights into the relationships between these
recombinant lineages. We compare the recombination events detected by 
sc2ts with the state-of-the-art RIPPLES method~\cite{Turakhia2022}, showing
that sc2ts is more sensitive and has a substantially lower false positive rate.
The output of sc2ts---incorporating
genealogical relationships, point mutations, deletions, and 
recombination---provides a basis for the study of SARS-CoV-2
that systematically accounts for all of these evolutionary forces.
This ARG, coupled with the rich suite of supporting software
in the tskit\cite{Ralph2020} and VCF-Zarr~\cite{Czech2025} ecosystems,
enables pandemic-scale analysis on a laptop. 
% Dunno about this! Feels important to finish on a wider note, helping
% the reader understand the significance of all of this?
These resources
will help us to learn from the vast troves of data collected during the
pandemic, and to prepare for the next.
